---
title: "Reproducible Research: Peer Assessment 1"
author: "Emilie Wolf"
date: "July 3, 2017"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r setup, results="hide", message=FALSE, echo=TRUE}
library(ggplot2)
library(dplyr)
unzip("activity.zip")
activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r histogram, echo=TRUE,message=FALSE,warning=FALSE}
## First let's aggregate the total steps per day using dplyr
totalsteps<- activity %>% group_by(date) %>% summarize(steps = sum(steps))

## Find the mean and median and assign to objects
m <- as.integer(mean(totalsteps$steps,na.rm = TRUE))
med <- median(totalsteps$steps,na.rm = TRUE)

## Create a histogram with two vertical lines representing mean and median.
h <- ggplot(totalsteps, aes(steps)) + 
        geom_histogram(bins = 21) + 
        ggtitle("Histogram of Daily Steps") +
        geom_vline(xintercept = m, color = "blue") +
        geom_vline(xintercept = med, color = "red") 
h
```

####The average number of steps per day was `r m` and the median was `r med`. 

Both the mean and median are shown on the histogram, but because they are very close together, they appear to make a solid line.

To show the different between a histogram and a barplot, let's make another graph for the total steps for each day, for fun!

####BONUS BARPLOT
```{r dailysteps, warning=FALSE, echo=TRUE}
g <- ggplot(totalsteps, aes(x=date,y=steps)) +
        geom_bar(stat = "identity") +  ##Must use stat="identity" to force 2 variables
        geom_hline(yintercept = m, color = "blue") +
        geom_hline(yintercept = med, color = "red") +
        theme(axis.text.x = element_text(angle=-90)) +
        ggtitle("Total Steps per Day, Oct-Nov 2012") + xlab("Date") + ylab("Steps")
g
```

I prefer the barplot when visualizing time-series data, but I see how the histogram helps in grouping if there is too much data.


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
